---
title: "Plotting time-course reassortment data"
author: "Guangchuang Yu, School of Basic Medical Sciences, Southern Medical University"
date: today
format:
  html:
    toc: true
    theme: none
    minimal: true
    embed-resources: true
vignette: >
  %\VignetteIndexEntry{Timeline workflows}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r style, echo=FALSE, results="asis", message=FALSE}
knitr::opts_chunk$set(tidy = FALSE,
                   message = FALSE)

r_libs <- unique(unlist(strsplit(
    paste(Sys.getenv(c("R_LIBS", "R_LIBS_USER", "R_LIBS_SITE")),
          collapse = .Platform$path.sep),
    .Platform$path.sep,
    fixed = TRUE
)))
r_libs <- r_libs[nzchar(r_libs)]
if (length(r_libs) > 0) {
    .libPaths(unique(c(r_libs, .libPaths())))
}
```

```{r echo=FALSE, results="hide", message=FALSE}
library(seqcombo)
library(ggplot2)
```


In a typical reassortment study you know **when** each virus was sampled, but
you rarely know where it should sit on the canvas. This vignette shows the full
path from two plain tables (one row per virus-segment record, one row per
reassortment flow) to a publication-ready figure with time on the x-axis.


## The shortest path

If you only want a quick look at what a timeline figure feels like,
`example_seqcombo_data("timeline")` returns ready-to-plot demo data:

```{r fig.width=9, fig.height=5}
demo <- example_seqcombo_data("timeline")
autoplot(demo, v_color = ~Host, v_fill = ~Host, link_style = "curve")
```

Time runs along the x-axis and `layout_timeline()` arranges the y positions so
that connected viruses stay close together. The rest of this document rebuilds
this kind of figure from scratch.

## Step 1: prepare long-format tables

The first table stores segment records. Each row is one *virus-segment*
combination:

```{r}
segment_df <- data.frame(
    id = rep(c("avian_1990", "human_1990", "swine_2000", "human_2009"), each = 8),
    sample_time = rep(c(1990, 1990, 2000, 2009), each = 8),
    segment = rep(c("PB2", "PB1", "PA", "HA", "NP", "NA", "M", "NS"), 4),
    color = c(
        rep("purple", 8),
        rep("red", 8),
        c("darkgreen", "darkgreen", "red", "darkgreen",
          "darkgreen", "purple", "red", "purple"),
        c("darkgreen", "lightgreen", "lightgreen", "darkgreen",
          "darkgreen", "purple", "red", "purple")
    ),
    host = rep(c("Avian", "Human", "Swine", "Human"), each = 8),
    stringsAsFactors = FALSE
)
head(segment_df)
```

The second table stores reassortment flows. `from` and `to` refer to virus ids;
the optional `segment` column records which segments were donated:

```{r}
flow_df <- data.frame(
    from = c("avian_1990", "human_1990", "swine_2000"),
    to = c("swine_2000", "swine_2000", "human_2009"),
    segment = c("HA", "NA", "PB1"),
    stringsAsFactors = FALSE
)
```

## Step 2: build `virus_info` and `flow_info`

At this stage there is no `y` column yet — only sampling time, which we hand to
the `x` argument as a column name:

```{r}
virus_info <- build_virus_info_from_long(segment_df, x = "sample_time",
                                         keep = "host")

flow_info <- build_flow_info_from_long(flow_df, segment = "segment")

virus_info[, c("id", "x")]
```

## Step 3: arrange the timeline

`layout_timeline()` keeps the time axis untouched (`time_col = "x"`) and fills
in sensible y positions based on the flow network:

```{r}
virus_info <- layout_timeline(virus_info, flow_info, time_col = "x")
```

## Step 4: plot

Wrap everything into a `seqcombo_data` object and call `autoplot()`:

```{r fig.width=9, fig.height=6}
data <- as_seqcombo_data(virus_info, flow_info)

autoplot(data, v_color = ~host, v_fill = ~host, link_style = "curve")
```

That is the complete workflow: two tables in, one figure out.

## Polishing the figure

Host colors work better with an explicit palette. The built-in
`scale_seqcombo_host()` wires both `colour` and `fill` aesthetics to a curated
host palette, and `build_segment_caption()` turns segment names into a readable
caption:

```{r fig.width=9, fig.height=6}
autoplot(data, v_color = ~host, v_fill = ~host, link_style = "curve") +
    scale_seqcombo_host() +
    labs(title = "Reassortment events leading to the 2009 H1N1 pandemic",
         caption = build_segment_caption(c("PB2", "PB1", "PA", "HA",
                                           "NP", "NA", "M", "NS"))) +
    scale_x_continuous(breaks = c(1990, 2000, 2009)) +
    theme_seqcombo_pub()
```

Labels can be added by attaching `label` and `label_position` columns before
building the plot:

```{r}
labels <- c("Avian H1N1", "Human seasonal H3N2", "Triple-reassortant swine H3N2",
            "2009 pandemic H1N1")

info_labeled <- data$virus_info
info_labeled$label <- labels
info_labeled$label_position <- c("left", "left", "below", "below")

data_labeled <- as_seqcombo_data(info_labeled, data$flow_info)
```

```{r fig.width=10, fig.height=6}
autoplot(data_labeled, v_color = ~host, v_fill = ~host,
         link_style = "curve") + theme_seqcombo_pub()
```

For dense timelines, elbow links often read more clearly than curves:

```{r fig.width=9, fig.height=6}
autoplot(data_labeled, v_color = ~host, v_fill = ~host,
         link_style = "elbow", link_elbow_position = 0.4) +
    theme_seqcombo_pub()
```

If both input tables carry a grouping variable, `facet_by` splits the history
into panels — useful when several lineages co-circulate:

```{r fig.width=10, fig.height=5}
info_facet <- data$virus_info
info_facet$Lineage <- c("North America", "North America",
                        "North America", "North America")
flow_facet <- data$flow_info
flow_facet$Lineage <- info_facet$Lineage[match(flow_facet$to, info_facet$id)]

autoplot(as_seqcombo_data(info_facet, flow_facet),
         v_color = ~host, v_fill = ~host,
         link_style = "curve", facet_by = "Lineage") +
    theme_seqcombo_pub()
```

## Notes

- Sampling times do not need to be years; dates (e.g., `"2024-03-01"`) or any
  numeric encoding works, since the values are passed through to ggplot2.
- If you prefer manual control at every step, replace `as_seqcombo_data()` +
  `autoplot()` with direct calls to `hybrid_plot()` — see the main vignette.
